---
title: "IML Presentation"
author: "Group 3"
format: html

execute: 
  echo: false
  warning: false
  message: false
---


```{r}
library(tidyverse) 
library(randomForest)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(plotly)
library(tidymodels)
library(ggthemes)
library(bbplot)
```




```{r}

set.seed(1148)
water <- read_csv("data/water_train.csv") |> 
  mutate(status_id = factor(status_id)) 
water_test <- read_csv("data/water_test.csv") 

```

```{r}
# Sample data for demonstration
# You should replace this with your own dataset
# Example data with latitude and longitude


my_sf <- read_sf("data/Africa_Boundaries.shp")

ggplot(my_sf) +
  geom_sf(fill = "cornsilk", color = "black") +
  geom_point(data = water, aes(x = lon, y = lat,color = status_id,shape = status_id), size = 3,alpha=0.2) +
  theme_minimal() +
  theme_void() #+ coord_sf(xlim = c(14.43, 35.45), ylim = c(2.22, 20))

```
# What amount of population has benefitted from the water technology installations ?

- The median local population within 1 km and water point populations for hand pumps and public tapstands dropped after 2018 while it was on the rise for motorised pumps for the same period.

- Does that mean the installations are happening away from the cities where the general population is much lower ?



```{r}
reach <- water %>% group_by(year(report_date),water_tech_category,status_id) %>% summarise("Median water point population" = median(water_point_population), "Median local population (1 km)" = median(local_population_1km)) %>% 
  pivot_longer(cols = "Median water point population":"Median local population (1 km)",names_to = "population_type",values_to = "population_reach")

pl <- ggplot(data = reach,
       aes(x = factor(`year(report_date)`), y = population_reach, color = status_id,group = status_id)) + geom_line(alpha =
                                                                                            0.4) + geom_point() + facet_wrap(population_type ~ water_tech_category) + theme_minimal() + theme(axis.text.x = element_text(angle = 45),plot.background = element_rect(fill = "cornsilk"),plot.title = element_text(hjust = 0.5),legend.text = element_text(size = 6),legend.title = element_text(size = 7)) +
  scale_color_manual(values = c("y" = "steelblue", "n" = "darkred")) + labs(x = "Report Year",y = "Population reach",title = "What amount of the population has benefited \n from the water technology installations?",caption = "Source: Water point data exchange (Wpdx)",color = "Water presence")

pl
```
# Are the recent installations moving away from the cities and focussing on regional areas ?

- For hand-pumps and public tapstands, the median distance from the cities have increased post 2019. __ This indicates that the recent installations are moving away from the cities, possibly targetting the regional areas.__

- On the other hand, we see that the median distance for motorised pumps lie around the 25 km mark for the recent as well as the previous installations. __This indicates that the recent motorised pump installations are still close to the cities__, due to which the water point and local population within 1 km is on the rise.

```{r}

water_tech_long <- water %>% mutate(period = if_else(year(report_date) >=2019,"Post 2019","Pre 2019")) %>% pivot_longer( 
                                          cols = distance_to_primary_road:distance_to_town,
                                          names_to = "distance_type",
                                          values_to = "distance") %>% select(c(water_tech_category,distance_type,distance,period))

water_tech_clean <- water_tech_long %>% mutate(distance_type_clean = case_when(distance_type == "distance_to_primary_road" ~ "Distance to primary road",
                                               distance_type == "distance_to_secondary_road" ~ "Distance to secondary road",
                                               distance_type == "distance_to_tertiary_road" ~ "Distance to tertiary road",
                                               distance_type == "distance_to_city" ~ "Distance to city",
                                               distance_type == "distance_to_town" ~ "Distance to town"))

ggplot(data = water_tech_clean,aes(y = distance_type_clean , x = distance/1000,color = period)) + geom_boxplot() + facet_grid(period~water_tech_category) + labs(title = "Distribution of distances from the water source \n pre and post 2019",x = "Distance (in Km)",y = "Type of distance measured",caption = "Source: Water point data exchange (Wpdx)") +  theme_minimal() + theme(axis.text.x = element_text(angle = 45),plot.background = element_rect(fill = "cornsilk"),legend.position = "none") + scale_color_canva(palette = "Summer inspired")
```

